
---
title: "Analyzing Student Burnout Using dplyr"
author: "Theresa Benny"
output: html_document
---

## Overview

This vignette demonstrates how to use dplyr functions from the tidyverse to analyze student mental health and burnout data. The goal is to explore patterns in burnout among students using a real dataset.

## Load libraries

```{r}
library(tidyverse)

## Load dataset
data <- read_csv("student_mental_health_burnout.csv")

data_small <- data %>%
  select(
    academic_pressure_score,
    burnout_level,
    daily_study_hours,
    daily_sleep_hours
  )

data_small <- data_small %>%
  mutate(
    pressure_group = case_when(
      academic_pressure_score <= 3 ~ "Low",
      academic_pressure_score <= 7 ~ "Medium",
      TRUE ~ "High"
    )
  )

data_small <- data_small %>%
  mutate(
    burnout_numeric = case_when(
      burnout_level == "Low" ~ 1,
      burnout_level == "Medium" ~ 2,
      burnout_level == "High" ~ 3
    )
  )

table(data_small$pressure_group)
```
```{r}
#Now that we have the select columns needed to analyze the statistics

burnout_group_analysis <- data_small %>%
  group_by(pressure_group) %>%
  summarise(
    avg_burnout = mean(burnout_numeric, na.rm = TRUE),
    student_count = n()
  ) %>%
  arrange(desc(avg_burnout))

burnout_group_analysis
```
Let's visualize this more clearly

```{r}
ggplot(burnout_group_analysis, aes(x = pressure_group, y = avg_burnout)) +
  geom_col() +
  labs(
    title = "Average Burnout by Academic Pressure Level",
    x = "Academic Pressure Level",
    y = "Average Burnout (1 = Low, 3 = High)"
  )
```
The analysis shows that average burnout levels are relatively consistent across different levels of academic pressure. 
While it might be expected that higher academic pressure leads to significantly higher burnout, this dataset does not show a strong or meaningful difference in average burnout across groups.

This demonstrates that not all variables have a strong relationship, and highlights the importance of validating assumptions with data. 
The tidyverse tools used in this example—such as dplyr for grouping and summarizing—help make it easy to explore and evaluate these relationships.
