---
title: "TidyVerse Vignette: purrr and dplyr"
author: "Ciara Bonnett"
format: html
editor: visual
---
## Introduction 
This vignette demonstrated how to use the `dplyr` and `purrr` packages within the TidyVerse. We are using the built-in `msleep` dataset, which contains information about the sleep patterns of mammals. This collaborative example is designed to show how we can transform raw data into a clean, summarized format. 


## Capability 1: Data Manipulation with dplyr

Here I am using`dplyr` to isolate herbivores. We calculate a new column for awake time and sort the results to see which heribivores sleep the most. This demonstrates the power of piping functions like `filter()`, `mutate(`, and `select()` together. 
```{r}
#| message: false
#| warning: false
library(tidyverse)

# Cleaning and transforming the mammal sleep data
herbivore_sleep <- msleep %>%
  filter(vore == "herbi") %>%
  mutate(awake_time = 24 - sleep_total) %>%
  select(name, sleep_total, awake_time, brainwt) %>%
  arrange(desc(sleep_total))
  
# Display the first few rows
knitr::kable(head(herbivore_sleep), caption = "Top Sleeping Herbivores")
```
## Capability 2: Functional Programming with purrr

The `purrr` package is used here to apply the `mean` function across every numeric column in the dataset at once. This is much more efficient than writing individual lines of code for every column. We use `map_df` to ensure the output remains in a clean data frame format. 
```{r}
#| message: false
#| warning: false
library(tidyverse)

# Use ggplot2::msleep to ensure R always sees it as the dataset
ggplot2::msleep %>%
  dplyr::select(where(is.numeric)) %>%
  purrr::map_df(~mean(.x, na.rm = TRUE)) %>%
  knitr::kable(caption = "Dataset Averages using purrr")
```
## Conclusion
This example shows how `dplyr` pipes (`%>%`) allows us to clean and transform data efficiently before passing it directly into a visualization.


## AI Transcript
Phase 1: Dataset Selection and Reasoning

User Inquiry: "So what dataset from kaggle do I need to use?"
AI Response & Logic: The AI suggested that while Kaggle is an option, using a built-in TidyVerse dataset like msleep is superior for a "collaborative vignette" because it ensures the code will run for all classmates instantly without requiring them to download external files or handle broken file paths.
Decision: The user chose to stay with the built-in msleep dataset for reliability and ease of extension by peers.

Phase 2: Troubleshooting the "Select" Error

User Inquiry: "Error in UseMethod('select') : no applicable method for 'select' applied to an object of class 'function'"
Technical Logic: * The Problem: The AI identified that the R environment was "masking" the TidyVerse select fuPhase 1: Dataset Selection and Reasoning

Phase 3: Final Polishing for Collaboration

User Inquiry: "Have I completed my code completely?"
AI Response & Logic: The AI confirmed the code was functional but suggested adding "Annotations" (descriptive text). This is a requirement for a "vignette" to ensure that the classmate who has to "extend" the code later understands the logic.


## Extension

To extend this Vignette I will show graphing functionality with `ggplot`.\
First we can import the library.\

```{r}
library(ggplot2)
```

The library ggplot allows us to create plots using our data. These plots can take many different forms and display many different types of data.\
In this example I will demonstrate a plot using techniques shown above.\
I will show the the body weight of animals versus their sleep total.

```{r}
# First create data 
sleep_bodyweight_df <- msleep %>%
  select(name, vore, sleep_total, bodywt)
  
# Specify the dataframe as data, mapping as an aes which contains the axes  
ggplot(data = sleep_bodyweight_df, mapping = aes(x = bodywt, y = sleep_total)) +
  geom_point() +
  labs( # Add labels
  title = "Bodyweight vs Sleep total of animals",
  x = "Body weight(kg)",
  y = "Total Sleep(hours)" # Rename axes with full names and units
)
```

We can see in the example above that values above 1000 are outliers that are making the rest of the graph tight.\
We can filter the data to remove these outliers as well as color the points based on their vore status.\
Just by adding this we can see new patterns like how no insects or omnivores in this data set go over 125 kg

```{r}
# Filter sleep_bodyweight_df 
sleep_bodyweight_df <- sleep_bodyweight_df %>%
  filter(bodywt < 1000)
  
# By adding in fill to our mapping we can get a graph that shows the vore aswell
ggplot(data = sleep_bodyweight_df, mapping = aes(x = bodywt, y = sleep_total, color = vore)) +
  geom_point() +
  labs( # Add labels
  title = "Bodyweight(less than 1000kg) vs Sleep total of animals",
  x = "Body weight(kg)",
  y = "Total Sleep(hours)" # Rename axes with full names and units
)
```

In conclusion we were able to see how we could use ggplot to create visualizations of data.


