---
title: "Tidyverse Vignette: Dplyr, Stringr & Ggplot2: Bechdel Test Dataset"
author: "Emily El Mouaquite"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The data used is the bechdel dataset from fivethirtyeight's library. The Bechdel Test measures representation for women in film. The three criteria of the test are: 

1. The movie has at least two female characters that have names.
2. The women in the movie talk to each other. 
3. They talk about something other than a man.

(https://bechdeltest.com/)

The dataset contains information on movies from 1970 - 2013, whether they passed or failed the Bechdel Test, and their budgets/ how much they grossed. 

Tidyverse, and its packages (specifically dplyr, ggplot2 and stringr) can be used to analyze and manipulate this data to explore Bechdel Test outcomes. 
 
```{r}
# Load necessary libraries
library(tidyverse)
library(fivethirtyeight) #install package if not already done so 
#load dataset
data("bechdel")
```

## Dplyr
Dplyr can be used to take a look at the data, as well as subset and manipulate it.
```{r}
# View the columns and some observations
glimpse(bechdel)
# Subset data based on movies that have passed the test
bechdel_pass <- bechdel %>%
                  filter(binary == "PASS")
#Add new variables 
bechdel %>%
  mutate(profit = domgross - budget)
#Sort data (highest to lowest grossing movies)
bechdel %>%
  arrange(desc(domgross))
#Group data to organize it (by year and why a movie passed or failed the test)
bechdel_why <- bechdel %>%
                  group_by(year, clean_test)
# Calculate the number of movies in each year and Bechdel test category group where each group represents a combination of year and clean_test outcome
summary_why <- bechdel_why %>%
  summarize(n = n())
```

## Ggplot2
Tidyverse's Ggplot2 can be used to create a line plot depicting the trends in the reasons why movies passed/ failed the Bechdel Test over time. 
```{r}
ggplot(summary_why, aes(x = year, y = n, color = clean_test)) +
  geom_line() +
  labs(
    title = "Bechdel Test Outcomes Over Time",
    x = "Year",
    y = "Number of Movies",
    color = "Test Result"
  )
```

## Stringr
Additionally, since much of this dataset does include text, stringr (along with dplyr) can be used to analuze movie titles and explore how certain words might relate to Bechdel Test outcomes.
```{r}
bechdel %>%
  mutate(woman_related_title = str_detect(title, regex("girl|woman|she|her", ignore_case = TRUE))) %>%
  group_by(binary, woman_related_title) %>%
  summarize(n = n(), .groups = "drop")

```

