---
title: "Vignette"
author: "Guibril Ramde"
format: html
editor: visual
---

## Approach

For this vignette assignment, I selected a dataset from Kaggle related to AI job market trends. The dataset was uploaded to my GitHub fork, and a raw file link was used to directly load the data into R.

The dataset is already well-structured and follows tidy data principles, with consistent variable names and appropriate data types. As a result, minimal data cleaning was required.

The analysis focuses on using Tidyverse packages, particularly `dplyr` and `ggplot2`, to transform, summarize, and visualize key insights such as skill demand, salary trends, and hiring patterns across countries.

## Code and Analyzation

```{r}
library(tidyverse)



url <- "https://raw.githubusercontent.com/japhet125/SPRING2026TIDYVERSE/refs/heads/main/AI_Job_Market_Trends_2026.csv"

load_jobs_data <- read_csv(url)
glimpse(load_jobs_data)
```

Column names

```{r}

colnames(load_jobs_data)
```

##Skills in demand

```{r}
skill_demand <- load_jobs_data |>
  summarise(
    python = mean(skills_python, na.rm = TRUE),
    sql = mean(skills_sql, na.rm = TRUE),
    ml = mean(skills_ml, na.rm = TRUE),
    dl = mean(skills_deep_learning, na.rm = TRUE),
    cloud = mean(skills_cloud, na.rm = TRUE)
  )
skill_demand
```

**##Salary by experience level**

```{r}
load_jobs_data |>
  group_by(experience_level) |>
  summarise(
    avg_salary = mean(salary, na.rm = TRUE)
    
     
  ) |>
  arrange(avg_salary) |>
  ggplot( aes(x = reorder(experience_level, avg_salary), y = avg_salary, fill = experience_level)) +
  geom_col() +
  labs(
    title = "Experience Level VS Average Salary",
    x = "Experience Level",
    y = "Average Salary"
  ) 


```

**##Top countries hiring**

```{r}
load_jobs_data |>
  count(country, sort = TRUE) |>
  slice_head(n =10) |>
  ggplot(aes(x = reorder(country, n), y =n, fill = country)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Countries Hiring for AI jobs",
    x = "Country",
    y = "Number of Job Listings"
  )
```

**##Conclusion**

This analysis demonstrates how Tidyverse tools can be effectively used to explore real-world datasets. The results show that skills such as Python, SQL, and machine learning are in high demand across AI-related roles. Additionally, salary trends vary significantly by experience level, with senior roles earning higher average salaries.

The analysis also highlights global hiring patterns, with certain countries showing higher demand for AI professionals. Overall, this vignette illustrates the power of Tidyverse for transforming, summarizing, and visualizing data to generate meaningful insights.


Extension by Desiree Thomas

I am extending Guibril's repo by using the Tidyverse to focus on data reshaping and categorical data management. With Guibril's code as it currently is, it is placed into a wide format which adds difficulty in being able to compare all of the skills at the same time and/or create visualizations. 

For this part, we are going to use the tidyr function to pivot the skills column into a Long format instead of wide format which will allow us to compare the skills in a single environment. Note that I am using |> instead of %>% as Guibril used them in his Tidyverse CREATE vignette. The boxplots are similar to what I used for my very first assignment in this class – it ended up being a nice callback. Stringr was used to clean up any names that may need extra attention. For the visualizations, I ensured that I added units such as $ and % to the chart for readability and I binned based on experience level.

```{r}
#Desiree's code extension

#Reshaping and Categorical Analysis


skills_comparison <- load_jobs_data |>  
  select(starts_with("skills_")) |>
  pivot_longer(cols = everything(), 
               names_to = "skill", 
               values_to = "is_required") |>
  mutate(skill = str_remove(skill, "skills_")) |> # Cleaned up names using stringr
  group_by(skill) |>
  summarise(percentage = mean(is_required, na.rm = TRUE) * 100)

# Visualization for Skills
ggplot(skills_comparison, aes(x = reorder(skill, percentage), y = percentage, fill = skill)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Prevalence of Skills in AI Job Postings",
       subtitle = "Pivoted from wide to long format for easier comparison",
       x = "Skill",
       y = "Percentage of Listings (%)") #always add units 

# Forcats

# Order logically: Entry -> Mid -> Senior -> Executive. instead of alphabetically (also something I learned when doing this during the first assignment)

load_jobs_data |>
  mutate(experience_level = fct_relevel(experience_level, 
                                        "Entry", "Mid", "Senior", "Executive")) |>
  ggplot(aes(x = experience_level, y = salary, fill = experience_level)) +
  geom_boxplot(alpha = 0.7) +
  theme_light() +
  labs(title = "Salary Distribution by Experience Level",
       subtitle = "Forcats used for progression on the x-axis",
       x = "Experience Level",
       y = "Salary ($)")
```

Unsurprisingly, as experience level increases so does salary. As a point of interest, the skills featured in listings did not contain a massive skew and were nearly equal with the exceptions being python and deep learning. Cloud skills were listed in the highest percentage of listings.
