---
title: "TidyVerse CREATE"
author: "Kristoff Oliphant"
date: "2026-03-25"
output: html_document
---

## Introduction

In this vignette, I'll show how to use the 'dplyr' package to view sleep patterns with mammals. I plan to use the 'msleep' dataset to work with functions like filter and select.

```{r}
library(tidyverse)
```

```{r}
sleepy_herbivores <- msleep %>%
  filter(vore == "herbi", sleep_total > 12) %>%
  select(name, vore, sleep_total)
```

```{r}
sleepy_herbivores
```

## Conclusion

In this code I wanted to focus on the sleep of herbivores from the msleep dataset. I used filter to find the class of vores that aligned with 'herbi' or in other words a herbivore. I also wanted to focus on herbivores that slept 12 hours or more. I then used select to specifically gather three columns which includes the name of the herbivore, the class of vore they are, and then the total amount of sleep they get.

## Extension

In this part, I wanted to extend the analysis by looking deeper into the herbivores that sleep more than 12 hours. First, I sorted them to see which animals sleep the most. Then, I calculated the average sleep time. Finally, I created a simple visualization.

```{r}
sleepy_herbivores_sorted <- sleepy_herbivores %>%
  arrange(desc(sleep_total))

sleepy_herbivores_sorted
```

```{r}
sleepy_herbivores_summary <- sleepy_herbivores %>%
  summarise(average_sleep = mean(sleep_total))

sleepy_herbivores_summary
```

```{r}
ggplot(sleepy_herbivores_sorted, aes(x = reorder(name, sleep_total), y = sleep_total)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Sleepiest Herbivores",
    x = "Animal",
    y = "Sleep Hours"
  )
```


