---
title: "Exploring College Majors with dplyr and ggplot2"
author: "Leon"
date: 
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
---


```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Load directly from fivethirtyeight's GitHub
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/recent-grads.csv"
majors <- read_csv(url)
```

---


```{r}
glimpse(majors)
```


```{r}
majors_clean <- majors %>%
  select(Major, Major_category, Total, Employed, Unemployed,
         Median, P25th, P75th, ShareWomen)

head(majors_clean)
```


```{r}
majors_clean <- majors_clean %>%
  mutate(
    unemployment_rate = Unemployed / (Employed + Unemployed),
    earnings_spread   = P75th - P25th
  )

head(majors_clean %>% select(Major, unemployment_rate, earnings_spread))
```

```{r}
avg_median <- mean(majors_clean$Median, na.rm = TRUE)

high_earners <- majors_clean %>%
  filter(Median > avg_median)

nrow(high_earners)
```


```{r}
category_summary <- majors_clean %>%
  group_by(Major_category) %>%
  summarise(
    n_majors          = n(),
    median_earnings   = median(Median, na.rm = TRUE),
    avg_unemployment  = mean(unemployment_rate, na.rm = TRUE),
    avg_share_women   = mean(ShareWomen, na.rm = TRUE)
  ) %>%
  arrange(desc(median_earnings))

category_summary
```

```{r}
majors_clean %>%
  arrange(desc(Median)) %>%
  select(Major, Major_category, Median) %>%
  slice_head(n = 10)
```

```{r fig.height=6, fig.width=8}
category_summary %>%
  mutate(Major_category = fct_reorder(Major_category, median_earnings)) %>%
  ggplot(aes(x = median_earnings, y = Major_category)) +
  geom_col(fill = "#2C7BB6") +
  geom_text(aes(label = scales::dollar(median_earnings)),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::dollar,
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Median Earnings by College Major Category",
    subtitle = "Recent graduates | FiveThirtyEight dataset",
    x        = "Median Annual Earnings",
    y        = NULL,
    caption  = "Source: fivethirtyeight/data/college-majors"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

unemployment rate vs median earnings


```{r fig.height=5, fig.width=8, warning=FALSE}
majors_clean %>%
  ggplot(aes(x = Median, y = unemployment_rate, color = Major_category)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Unemployment Rate vs. Median Earnings by Major",
    x        = "Median Annual Earnings",
    y        = "Unemployment Rate",
    color    = "Category",
    caption  = "Source: fivethirtyeight/data/college-majors"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold"),
    legend.position = "bottom",
    legend.text  = element_text(size = 7)
  ) +
  guides(color = guide_legend(nrow = 4))
```


| Function | Package | Purpose |
|---|---|---|
| `read_csv()` | `readr` | Load CSV data from URL |
| `glimpse()` | `dplyr` | Quick data overview |
| `select()` | `dplyr` | Keep specific columns |
| `mutate()` | `dplyr` | Create new variables |
| `filter()` | `dplyr` | Subset rows by condition |
| `group_by()` | `dplyr` | Group for aggregation |
| `summarise()` | `dplyr` | Compute group-level stats |
| `arrange()` | `dplyr` | Sort rows |
| `fct_reorder()` | `forcats` | Reorder factor levels for plotting |
| `ggplot()` + `geom_col()` / `geom_point()` | `ggplot2` | Data visualization |
