---
title: "Tidyverse Create"
author: "Long Lin"
format: html
editor: visual
---

## Introduction

This vignette shows how to use Tidyverse packages like `dplyr` and `ggplot2` to organize data and display findings in a chart. The chosen data source named `global grocery inflation` is from Kaggle.

Source: https://www.kaggle.com/datasets/waddahali/global-grocery-inflation-20252026/data?select=breakfast+basket.csv

## Reading in the data with Tidyverse

We read in the data from Kaggle by first uploading the data to github and then using the `read_csv` method within the `tidyverse` library on a raw github link. Then we use the `glimpse` method within `dplyr` to visualize the columns within the data set.

```{r}
library(tidyverse)

url <- "https://raw.githubusercontent.com/longflin/DATA-607/refs/heads/main/TIDYVERSE2026/GlobalGroceryInflation.csv"

grocery_inflation <- read_csv(
  file = url,
  show_col_types = FALSE,
  progress = FALSE
)

glimpse(grocery_inflation)

```
Next, we use `select` within `dplyr` to select specific columns of interest. For this vignette, we will focus on the price of milk in target cities such as `New York`, `Los Angeles`, `London`, `Paris`, `Beijing`, and `Tokyo`.  In order to do this, we use `filter` for `Item` that is equal to `Milk (1 Liter)` and for `City` that are in the `target_cities` variable.

```{r}
target_cities <- c("New York", "Los Angeles", "London", "Paris", "Beijing", "Tokyo")

milk_inflation <- grocery_inflation %>%
  select(City, Country, Item, Month, YoY_Inflation_Estimate_Pct) %>%
  filter(Item == "Milk (1 Liter)") %>%
  filter(City %in% target_cities)

head(milk_inflation)
```
Here, we use `summarize` to create a new column named `Average_YoY` to grab the average of all the months for each city.

```{r}
milk_inflation_group <- milk_inflation %>%
  group_by(City) %>% 
  summarize(
    Average_YoY = mean(YoY_Inflation_Estimate_Pct, na.rm = TRUE)
  )

head(milk_inflation_group)
```
In order to visualize the data, we use `ggplot2` to create a bar chart to display the results.

```{r}
library(ggplot2)

ggplot(milk_inflation_group, aes(x = reorder(City, Average_YoY), 
                                       y = Average_YoY, 
                                       fill = City)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(Average_YoY, 1)), 
            hjust = -0.2, 
            size = 3.5) + 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  labs(
    title = "Average Milk Inflation by City for Oct 2025 to March 2026",
    x = "City",
    y = "Average Year over Year Inflation (%)"
  ) +
  theme_minimal()
```
## Conclusion
Tidyverse is a very powerful tool for managing and transforming data.  It allows you to create easy to read charts created from a data source of your choice.  In this example, packages such as `dplyr` and `ggplot2`, made it very easy to manipulate the data and display the results.