---
title: "OnyVignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OnyVignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(OnyVignette)
```

This vignette is a short summary of the 'DT' package, specifically the function 'datatable()', and how it can be used to create interactive data tables on HTML pages with a given input. See the official documentation here. Some added interaction capabilities are sorting, filtering and searching.

Example 1: Basic Usage

The dataframe() function will turn any dataframe or matrix to an interactive table. This is demonstrated on a synthesized student mental health dataset from kaggle in the code chunks below:

```{r - Install Package}

install.packages('DT')

library('DT')
```


```{r - Load Data}

student_frame <- data.frame(read.csv("https://raw.githubusercontent.com/ifi-ony/week9/refs/heads/main/studentframe.csv"))
```

```{r - Call Datatable}

datatable(student_frame)
```

Example 2: Customization (Options Argument)

The table's features can be customized using the 'options' argument of the datatable function:

```{r - Options}

datatable(student_frame,
          options = list( #takes list as input
          scrollX = TRUE, # adds horizontal scroll for user
          pagelength = 20 # shows 20 observation per page instead of default 10
          
          ))
```

Many more capabilities can be explored such as button extensions and column/row formatting. This will be explored in an extended version of this vignette
