---
title: "TidyVerse CREATE: Exploring College Major Earnings with dplyr, ggplot2, and tidyr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This vignette demonstrates core capabilities of three TidyVerse packages: **dplyr**,
**ggplot2**, and **tidyr**, using the `college_recent_grads` dataset from the
`fivethirtyeight` package.

Note: to install fivethirtyeight in the console: `install.packages("fivethirtyeight")`

## Load Libraries and Data

```{r load}
library(tidyverse)
library(fivethirtyeight)

data("college_recent_grads")
glimpse(college_recent_grads)
```

---

## Part 1: dplyr -- Selecting, Filtering, and Summarizing

### select() -- Pick the columns you need

`select()` lets you keep only the columns relevant to your analysis, which makes
data frames easier to read and work with.

```{r select}
major_earnings <- college_recent_grads |>
  select(major, major_category, total, median, p25th, p75th, unemployment_rate)

head(major_earnings)
```

### filter() -- Subset rows by condition

`filter()` keeps only the rows that satisfy a logical condition. Here we pull out
majors where the median salary exceeds $50,000.

```{r filter}
high_earning <- major_earnings |>
  filter(median > 50000)

high_earning |>
  select(major, major_category, median) |>
  arrange(desc(median))
```

### mutate() -- Create new columns

`mutate()` adds new columns that are functions of existing ones. We can compute an
interquartile range (IQR) of earnings to measure salary spread within each major.

```{r mutate}
major_earnings <- major_earnings |>
  mutate(iqr_earnings = p75th - p25th)

major_earnings |>
  select(major, median, iqr_earnings) |>
  arrange(desc(iqr_earnings)) |>
  head(10)
```

### group_by() + summarize() -- Aggregate by category

`group_by()` splits the data by a categorical variable, and `summarize()` computes
summary statistics for each group. This is one of the most powerful patterns in dplyr.

```{r group-summarize}
category_summary <- major_earnings |>
  group_by(major_category) |>
  summarize(
    n_majors       = n(),
    median_salary  = median(median, na.rm = TRUE),
    mean_unemp     = mean(unemployment_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(median_salary))

category_summary
```

### arrange() -- Sort your results

We already used `arrange()` above, but it deserves its own callout. Sorting by a
computed summary makes it easy to spot the top and bottom categories at a glance.

```{r arrange}
category_summary |>
  arrange(mean_unemp)
```

Agriculture and Natural Resources has both a respectable median salary and the
lowest average unemployment rate among these categories.

---

## Part 2: ggplot2 -- Visualizing the Data

### Bar chart -- Median salary by major category

`geom_col()` draws bars whose heights correspond to values in the data (as opposed
to `geom_bar()`, which counts rows). `coord_flip()` makes long category labels
readable.

```{r bar-chart, fig.width=8, fig.height=6}
ggplot(category_summary, aes(x = reorder(major_category, median_salary),
                              y = median_salary)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Median Salary by Major Category",
    x = NULL,
    y = "Median Salary ($)"
  ) +
  theme_minimal()
```

### Scatter plot -- Salary vs. unemployment rate

Each point is one major. `geom_smooth()` overlays a LOESS trend line so we can see
whether higher-paying majors tend to have lower unemployment.

```{r scatter, fig.width=8, fig.height=6}
ggplot(major_earnings, aes(x = unemployment_rate, y = median)) +
  geom_point(aes(color = major_category), alpha = 0.7, size = 2) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "Median Salary vs. Unemployment Rate by Major",
    x = "Unemployment Rate",
    y = "Median Salary ($)",
    color = "Category"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7))
```

### Box plot -- Salary distributions within categories

Box plots reveal the spread and outliers within each category, not just the center.

```{r boxplot, fig.width=8, fig.height=6}
ggplot(major_earnings, aes(x = reorder(major_category, median, FUN = stats::median),
                            y = median)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red") +
  coord_flip() +
  labs(
    title = "Distribution of Median Salaries Within Each Major Category",
    x = NULL,
    y = "Median Salary ($)"
  ) +
  theme_minimal()
```

---

## Part 3: tidyr -- Reshaping Data

### pivot_longer() -- Wide to long format

Many ggplot geometries work best with "long" data, where each row is a single
observation. The salary percentile columns (`p25th`, `median`, `p75th`) are
currently spread across three columns. `pivot_longer()` stacks them into one
column so we can map the percentile label to color or facet.

```{r pivot-longer}
top5 <- major_earnings |>
  arrange(desc(median)) |>
  head(5)

top5_long <- top5 |>
  select(major, p25th, median, p75th) |>
  pivot_longer(
    cols      = c(p25th, median, p75th),
    names_to  = "percentile",
    values_to = "salary"
  )

top5_long
```

### Visualizing the reshaped data

Now that we have long-format data, a grouped bar chart comparing the three
percentiles for each major is straightforward.

```{r grouped-bar, fig.width=9, fig.height=5}
top5_long <- top5_long |>
  mutate(percentile = factor(percentile,
                             levels = c("p25th", "median", "p75th"),
                             labels = c("25th Pctile", "Median", "75th Pctile")))

ggplot(top5_long, aes(x = reorder(major, salary), y = salary, fill = percentile)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Salary Percentiles for Top 5 Earning Majors",
    x = NULL,
    y = "Salary ($)",
    fill = "Percentile"
  ) +
  theme_minimal()
```

### pivot_wider() -- Long back to wide

`pivot_wider()` is the inverse operation. If a downstream model or table expects
one column per percentile, you can spread the data back out.

```{r pivot-wider}
top5_wide <- top5_long |>
  pivot_wider(
    names_from  = percentile,
    values_from = salary
  )

top5_wide
```

---

## Conclusion

This vignette walked through the core verbs of three TidyVerse packages:

- **dplyr**: `select()`, `filter()`, `mutate()`, `group_by()`, `summarize()`, `arrange()`
- **ggplot2**: `geom_col()`, `geom_point()`, `geom_smooth()`, `geom_boxplot()`
- **tidyr**: `pivot_longer()`, `pivot_wider()`

Together these tools cover a typical analysis workflow: subset and transform the
data, compute summaries, reshape for plotting, and visualize the results.
