---
title: "Shawn_Ganz_Vignette"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

I'll be using the National Center for Health Statistics' (NCHS) National Health and Nutrition Examination Survey (NHANES) study from August 2021-August 2023.

- [dataset]("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.xpt")
- [dataset information](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.htm)
- [study information](https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Dietary&Cycle=2021-2023)

I'll show how simple it is by using tidyverse to download, prep, and display the data. We download the entire dataset, select the columns we want to showcase (DBD100, DR1DRSTZ), select, filter, and then display.

## Codebase

```{r}

library(tidyverse)
library(haven)
library(ggplot2)

df <- read_xpt("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.xpt") |>
  select(salt_added_table = DBD100,
         recall_status = DR1DRSTZ) |>
  filter(recall_status == 1, salt_added_table %in% c(1, 2, 3, 7)) |>
  mutate(
    salt_added_table = case_when(
      salt_added_table == 1 ~ "Rarely",
      salt_added_table == 2 ~ "Occasionally",
      salt_added_table == 3 ~ "Very often",
      salt_added_table == 7 ~ "Refused"
    )
  ) |>
  count(salt_added_table)

ggplot(df, aes(x = salt_added_table, y = n)) +
  geom_col() +
  labs(
    title = "How Often Salt Is Added to Food at the Table",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal()
```

## Conclusion

Just a quick example to show that power of tidyverse in altering a DF with 8860 entries and over 50 columns, to a dataframe with two columns that are altered for whatever you need - in this case showing how often Salt is Added to food on the table.




## Extension by Long Lin

For the data and plot above, I chose to clean it up a little by using `factor` to reorder the chart in a more logical order rather than just alphabetical.  I also changed the color of the bars in order to improve readability by using a cleaner color.

```{r}
library(tidyverse)
library(haven)

df <- read_xpt("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.xpt") |>
  select(salt_added_table = DBD100,
         recall_status = DR1DRSTZ) |>
  filter(recall_status == 1, salt_added_table %in% c(1, 2, 3, 7)) |>
  mutate(
    # used factor() to change it to a more logical ordering
    salt_added_table = factor(
      salt_added_table, 
      levels = c(1, 2, 3, 7), 
      labels = c("Rarely", "Occasionally", "Very often", "Refused")
    )
  ) |>
  count(salt_added_table)

ggplot(df, aes(x = salt_added_table, y = n)) +
  geom_col(fill = "steelblue") + # Added a cleaner color
  labs(
    title = "How Often Salt Is Added to Food at the Table",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal()
```

Next I expanded on the data by grabbing the `sodium_mg` and `calories_kcal` variables as well from the dataset in order to normalize the data so that the salt intake is based on the caloric intake as well. I did this because someone who eats more is more likely to consume more salt along with the calories.

```{r}
library(tidyverse)
library(haven)

df <- read_xpt("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DR1TOT_L.xpt") |> 
  select(
    salt_added_table = DBD100,
    recall_status    = DR1DRSTZ,
    sodium_mg        = DR1TSODI,
    calories_kcal    = DR1TKCAL
  ) |> 
  filter(recall_status == 1, salt_added_table %in% c(1, 2, 3), calories_kcal > 0) |> 
  mutate(
    salt_added_table = factor(salt_added_table, 
                              levels = c(1, 2, 3), 
                              labels = c("Rarely", "Occasionally", "Very often")),
    
    sodium_density = (sodium_mg / calories_kcal) * 1000
  )

ggplot(df, aes(x = salt_added_table, y = sodium_density, fill = salt_added_table)) +
  geom_boxplot(outlier.alpha = 0.2) +
  coord_cartesian(ylim = c(0, 4000)) +
  labs(
    title = "Sodium Density by Table Salt Habit",
    x = "How often is salt added at the table?",
    y = "Sodium Density (mg / 1,000 kcal)",
    fill = "Response"
  ) +
  theme_minimal()
```
Based on the plot, it looks like whether people add salt to their food very often or rarely, they have a similar sodium density.  This means that the salt added at the table is negligible compared  to the salt already present in the food.
