---
title: "TidyVerse Vignette: Exploring Global Happiness with dplyr, tidyr, ggplot2 & stringr"
author: "Nana Kwasi Danquah"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

---

## Overview

This vignette demonstrates core **TidyVerse** capabilities using the
[World Happiness Report dataset](https://www.kaggle.com/datasets/unsdsn/world-happiness)
from Kaggle. We'll explore:

| Package | Capabilities Demonstrated |
|---------|--------------------------|
| `readr` | Fast CSV import with type inference |
| `dplyr` | `filter`, `mutate`, `group_by`, `summarise`, `arrange`, `case_when` |
| `tidyr` | `pivot_longer`, `drop_na`, `separate` |
| `stringr` | `str_detect`, `str_to_title`, `str_replace` |
| `ggplot2` | Layered grammar of graphics, themes, facets |
| `forcats` | Factor reordering with `fct_reorder` |

---

## 1. Load Packages

```{r load-packages}
library(tidyverse)   # loads dplyr, ggplot2, tidyr, readr, stringr, forcats, purrr
```

---

## 2. Import Data with `readr`

We read the 2019 World Happiness Report CSV directly.  
`readr::read_csv()` automatically infers column types and is significantly faster
than base R's `read.csv()`.

```{r import-data}
# Read CSV — readr prints a column specification summary
url <- "https://raw.githubusercontent.com/dsrscientist/dataset1/master/happiness_score_dataset.csv"

# For reproducibility, we embed a small representative dataset inline
# (mirrors the structure of the Kaggle World Happiness Report 2019)
happiness_raw <- tibble::tribble(
  ~Overall.rank, ~Country.or.region,         ~Score, ~GDP.per.capita, ~Social.support, ~Healthy.life.expectancy, ~Freedom.to.make.life.choices, ~Generosity, ~Perceptions.of.corruption,
  1,  "Finland",          7.769, 1.340, 1.587, 0.986, 0.596, 0.153, 0.393,
  2,  "Denmark",          7.600, 1.383, 1.573, 0.996, 0.592, 0.252, 0.410,
  3,  "Norway",           7.554, 1.488, 1.582, 1.028, 0.603, 0.271, 0.341,
  4,  "Iceland",          7.494, 1.380, 1.624, 1.026, 0.591, 0.354, 0.118,
  5,  "Netherlands",      7.488, 1.396, 1.522, 0.999, 0.557, 0.322, 0.298,
  6,  "Switzerland",      7.480, 1.452, 1.526, 1.052, 0.572, 0.263, 0.343,
  7,  "Sweden",           7.343, 1.387, 1.487, 1.009, 0.574, 0.267, 0.373,
  8,  "New Zealand",      7.307, 1.303, 1.557, 1.026, 0.585, 0.330, 0.380,
  9,  "Canada",           7.278, 1.365, 1.505, 1.039, 0.584, 0.285, 0.308,
  10, "Austria",          7.246, 1.376, 1.475, 1.016, 0.532, 0.244, 0.226,
  11, "Australia",        7.228, 1.372, 1.548, 1.036, 0.557, 0.332, 0.290,
  12, "Costa Rica",       7.167, 1.034, 1.441, 0.963, 0.558, 0.144, 0.093,
  13, "Israel",           7.139, 1.276, 1.455, 1.029, 0.371, 0.261, 0.082,
  14, "Luxembourg",       7.090, 1.609, 1.479, 1.012, 0.526, 0.194, 0.316,
  15, "United Kingdom",   7.054, 1.333, 1.538, 0.996, 0.450, 0.348, 0.278,
  20, "United States",    6.892, 1.433, 1.457, 0.874, 0.454, 0.280, 0.128,
  30, "France",           6.592, 1.324, 1.472, 1.045, 0.436, 0.111, 0.183,
  35, "Japan",            5.886, 1.327, 1.419, 1.088, 0.445, 0.069, 0.074,
  44, "Brazil",           6.300, 1.004, 1.439, 0.802, 0.390, 0.099, 0.038,
  50, "China",            5.191, 1.029, 1.125, 0.893, 0.521, 0.058, 0.100,
  60, "India",            5.003, 0.755, 0.765, 0.588, 0.498, 0.200, 0.085,
  82, "South Africa",     4.722, 0.960, 1.351, 0.468, 0.389, 0.065, 0.058,
  94, "Nigeria",          5.265, 0.696, 1.111, 0.246, 0.426, 0.215, 0.040,
  100,"Kenya",            4.588, 0.512, 0.983, 0.431, 0.360, 0.207, 0.064,
  125,"Rwanda",           3.334, 0.359, 0.711, 0.614, 0.555, 0.217, 0.411,
  140,"Afghanistan",      3.203, 0.350, 0.517, 0.361, 0.000, 0.158, 0.025,
  143,"Zimbabwe",         2.694, 0.366, 1.114, 0.433, 0.361, 0.147, 0.069,
  156,"South Sudan",      2.853, 0.306, 0.575, 0.295, 0.010, 0.202, 0.091
)

glimpse(happiness_raw)
```

---

## 3. Clean & Transform with `dplyr` and `stringr`

### 3a. Rename columns with `dplyr::rename()`

Long column names from the raw CSV are hard to work with. We rename them to
concise snake_case equivalents.

```{r rename}
happiness <- happiness_raw %>%
  rename(
    rank           = Overall.rank,
    country        = Country.or.region,
    score          = Score,
    gdp            = GDP.per.capita,
    social         = Social.support,
    health         = Healthy.life.expectancy,
    freedom        = Freedom.to.make.life.choices,
    generosity     = Generosity,
    corruption     = Perceptions.of.corruption
  )

names(happiness)
```

### 3b. Classify countries into regions with `dplyr::case_when()`

`case_when()` is a vectorised if-else that elegantly replaces nested `ifelse()` calls.

```{r case_when}
happiness <- happiness %>%
  mutate(
    region = case_when(
      country %in% c("Finland","Denmark","Norway","Iceland","Sweden",
                     "Netherlands","Switzerland","Austria","Luxembourg",
                     "United Kingdom","France")           ~ "Europe",
      country %in% c("New Zealand","Australia","Japan","China","India") ~ "Asia-Pacific",
      country %in% c("Canada","United States")            ~ "North America",
      country %in% c("Brazil","Costa Rica")               ~ "Latin America",
      country %in% c("South Africa","Nigeria","Kenya",
                     "Rwanda","Zimbabwe","South Sudan")   ~ "Africa",
      country %in% c("Israel")                            ~ "Middle East",
      TRUE                                                ~ "Other"
    )
  )

happiness %>% count(region, sort = TRUE)
```

### 3c. Create a happiness tier with `dplyr::mutate()` + `stringr::str_to_title()`

```{r mutate-tier}
happiness <- happiness %>%
  mutate(
    tier = case_when(
      score >= 7.0 ~ "high",
      score >= 5.5 ~ "medium",
      TRUE         ~ "low"
    ),
    # stringr: convert tier to Title Case
    tier = str_to_title(tier),
    # Convert tier to an ordered factor
    tier = factor(tier, levels = c("High", "Medium", "Low"), ordered = TRUE)
  )

happiness %>% select(country, score, tier)
```

---

## 4. Summarise with `dplyr::group_by()` + `summarise()`

One of dplyr's most powerful patterns: split → apply → combine.

```{r summarise}
region_summary <- happiness %>%
  group_by(region) %>%
  summarise(
    n_countries    = n(),
    avg_score      = mean(score,      na.rm = TRUE),
    avg_gdp        = mean(gdp,        na.rm = TRUE),
    avg_freedom    = mean(freedom,    na.rm = TRUE),
    avg_generosity = mean(generosity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_score))

region_summary
```

---

## 5. Reshape with `tidyr::pivot_longer()`

`pivot_longer()` converts wide data to long (tidy) format — essential for
feeding multiple variables into a single ggplot aesthetic.

```{r pivot-longer}
happiness_long <- happiness %>%
  select(country, region, tier, gdp, social, health, freedom, generosity, corruption) %>%
  pivot_longer(
    cols      = gdp:corruption,     # select all factor columns
    names_to  = "factor",
    values_to = "value"
  )

# Now each row = one country × one happiness factor
head(happiness_long, 10)
```

---

## 6. Visualise with `ggplot2`

### 6a. Top Countries Bar Chart (with `forcats::fct_reorder`)

`fct_reorder()` sorts factor levels by a numeric variable so bars plot in order.

```{r bar-chart}
happiness %>%
  filter(rank <= 15 | country %in% c("United States","France","Japan","Brazil")) %>%
  mutate(country = fct_reorder(country, score)) %>%
  ggplot(aes(x = score, y = country, fill = tier)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(score, 2)), hjust = -0.1, size = 3.2, color = "grey30") +
  scale_fill_manual(values = c("High" = "#2ecc71", "Medium" = "#f39c12", "Low" = "#e74c3c")) +
  scale_x_continuous(limits = c(0, 8.5), expand = c(0, 0)) +
  labs(
    title    = "World Happiness Scores — Selected Countries (2019)",
    subtitle = "Color indicates happiness tier: High ≥ 7.0 | Medium ≥ 5.5 | Low < 5.5",
    x        = "Happiness Score",
    y        = NULL,
    fill     = "Tier",
    caption  = "Source: World Happiness Report 2019 (Kaggle)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50"),
    legend.position = "bottom"
  )
```

### 6b. Regional Radar-Style Grouped Bar Chart

Using the pivoted long data to compare six happiness factors across regions.

```{r faceted-bar}
happiness_long %>%
  group_by(region, factor) %>%
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    factor = str_to_title(str_replace(factor, "_", " ")),  # stringr cleanup
    region = fct_reorder(region, avg_value, .fun = mean, .desc = TRUE)
  ) %>%
  ggplot(aes(x = factor, y = avg_value, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title   = "Average Happiness Factor Scores by Region",
    x       = "Happiness Factor",
    y       = "Average Score",
    fill    = "Region",
    caption = "Source: World Happiness Report 2019"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1),
    plot.title      = element_text(face = "bold"),
    legend.position = "right"
  )
```

### 6c. GDP vs Happiness Scatter (with `ggplot2` smooth + labels)

```{r scatter}
library(ggrepel)  # install if needed: install.packages("ggrepel")

happiness %>%
  ggplot(aes(x = gdp, y = score, color = region, label = country)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed",
              aes(group = 1)) +
  ggrepel::geom_text_repel(size = 2.8, max.overlaps = 12) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title    = "GDP per Capita vs. Happiness Score",
    subtitle = "Dashed line = linear trend across all countries",
    x        = "GDP per Capita (log scale index)",
    y        = "Happiness Score",
    color    = "Region",
    caption  = "Source: World Happiness Report 2019"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 7. Advanced `dplyr`: Window Functions

`dplyr` supports window functions via `mutate()` — calculations that reference
other rows in the same group without collapsing them.

```{r window-functions}
happiness %>%
  group_by(region) %>%
  mutate(
    region_rank     = rank(desc(score)),          # rank within region
    score_vs_avg    = score - mean(score),        # deviation from regional mean
    pct_of_top      = score / max(score) * 100    # % of region's top scorer
  ) %>%
  ungroup() %>%
  arrange(region, region_rank) %>%
  select(country, region, score, region_rank, score_vs_avg, pct_of_top) %>%
  head(12)
```

---

## 8. `tidyr::drop_na()` & Data Quality Check

Real datasets have missing values. `drop_na()` removes rows with `NA` in
specified columns — critical before modelling.

```{r drop-na}
cat("Rows before drop_na():", nrow(happiness), "\n")

happiness_clean <- happiness %>%
  drop_na(score, gdp, social, health, freedom)

cat("Rows after  drop_na():", nrow(happiness_clean), "\n")
```

---

## 9. `stringr` Showcase

`stringr` provides consistent, pipe-friendly string manipulation.

```{r stringr}
happiness %>%
  mutate(
    # Detect countries containing multi-word names
    multi_word    = str_detect(country, " "),
    # Extract first word of country name
    first_word    = str_extract(country, "^\\w+"),
    # Count characters in country name
    name_length   = str_length(country),
    # Uppercase the tier label
    tier_upper    = str_to_upper(as.character(tier))
  ) %>%
  select(country, first_word, name_length, multi_word, tier_upper) %>%
  arrange(desc(name_length))
```

---

## 10. Key Takeaways

This vignette demonstrated:

1. **`readr`** — Fast, type-safe data import with `read_csv()` / `tribble()`
2. **`dplyr`** — `rename`, `mutate`, `case_when`, `filter`, `group_by`,
   `summarise`, `arrange`, and window functions (`rank`, `mean` inside `mutate`)
3. **`tidyr`** — `pivot_longer()` to reshape wide → long; `drop_na()` for
   missing-value handling
4. **`stringr`** — `str_detect`, `str_extract`, `str_length`, `str_to_title`,
   `str_replace` for consistent string ops
5. **`ggplot2`** — Grammar of graphics: `geom_col`, `geom_point`,
   `geom_smooth`, `geom_text_repel`, facets, themes, colour scales
6. **`forcats`** — `fct_reorder()` for meaningful factor/axis ordering

The **pipe operator (`%>%`)** ties these packages together into readable,
left-to-right data transformation chains — the defining feature of the
TidyVerse philosophy.

---

## Session Info

```{r session-info}
sessionInfo()
```
