---
title: "Tidyverse Create"
author: "Sam Barbaro"
format: html
editor: visual
---

## Stringr Package

<https://stringr.tidyverse.org/>

Correct punctuation is very important to me, so I've decided to explore the Tidyverse's Stringr package, which allows you to manipulate strings.

I'm using a jobs data set from NYC open data to demonstrate some of the cleaning capabilities.

```{r}
library(tidyverse)

csv <- "https://raw.githubusercontent.com/samanthabarbaro/data607/refs/heads/main/nyc_jobs.csv"

jobs <- read.csv(csv, header = TRUE)

glimpse(jobs)
```

### str_replace

I don't like that the column names have periods instead of underscores (who would do that?) . Let's replace them with str_replace:

```{r}
jobs <- jobs %>%
  rename_with(~ str_replace(.x, "\\.", "_"))
```

### Changing case

The agencies are in all caps. Let's use the string to title function to change them to title case:

```{r}
jobs <- jobs |> 
  mutate(Agency = str_to_title(Agency))

```

Let's use str_to_lower on business title to make all the jobs lowercase.

```{r}
jobs <- jobs |> 
  mutate(Business_Title = str_to_lower(Business_Title))
```

You can also use str_to_upper for uppercase and str_to_sentence for title case.

```{r}
jobs <- jobs |> 
  mutate(Posting_Type = str_to_upper(Posting_Type)) |>
  mutate(Title_Classification = str_to_sentence(Title_Classification))
```

### str_count

str_count lets you count anything that includes a certain string. Some of these functions work as-is on vectors, but require "regex" when used with data frames:

```{r}
jobs_with_counts <- jobs |> 
    mutate(data_mentions = str_count(Business_Title, regex("data")))
```

### str_like

str_like - detect string matches in a vector or specified column. Here, we use it to create a new data frame with data jobs:

```{r}
data_jobs <- jobs |> 
  filter(str_detect(Business_Title, regex("data", ignore_case = TRUE)))
```

For two or more matches, use a pipe inside the quotes:

```{r}
legal_jobs <- jobs |> 
  filter(str_detect(Business_Title, regex("lawyer|attorney|paralegal", ignore_case = TRUE)))
```

### str_split

str_split breaks up a vector or column based on a specified character. We'll break the "Career_Level" column at a an open parentheses to get rid of extra information:

```{r}
jobs_split <- jobs |>
  mutate(seniority = str_split_i(Career_Level, " \\(", 1))
```

### str_order

Str_order is designed for vectors, but can be used as an alternative to dplyr's arrange:

```{r}
# Sorting legal_jobs by Business_Title alphabetically
ordered_indices <- str_order(legal_jobs$Business_Title)
legal_jobs_sorted <- legal_jobs[ordered_indices, ]

# with dplyr
legal_jobs_sorted_2 <- legal_jobs |> arrange(Agency)
```

### str_equal

str_equal can detect and subset exact matches:

```{r}
# Find jobs where the title is EXACTLY "Paralegal" (case sensitive)
agency_attorneys_only <- legal_jobs |> 
  filter(str_equal(Business_Title, "agency attorney"))
```
