---
title: 'TidyVerse Vignette: Visualizing Student Mental Health & Burnout with ggplot2'
author: "Mark Hamer"
date: "March 28, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What's This About?

In this vignette, we're going to use **ggplot2** to dig into a dataset about student mental health and burnout. The data comes from a student survey, so it's pretty relatable stuff — stress, sleep, burnout, whether people actually asked for help.

We'll walk through five different plot types, all built with ggplot2:

- `geom_histogram()` — to see how burned out students are overall
- `geom_boxplot()` — to compare burnout across genders
- `geom_point()` + `geom_smooth()` — to see if stress and burnout move together
- `geom_bar()` — to check if burnout gets worse over the years
- `facet_wrap()` — to split one plot into panels for easy comparison

Each plot comes with an explanation of what it's doing and what to look for.

---

## The Dataset

The data is from Kaggle:
[Student Mental Health and Burnout](https://www.kaggle.com/datasets/sharmajicoder/student-mental-health-and-burnout)

Here's a quick rundown of the columns we'll be working with:

| Variable | Description |
|---|---|
| `Age` | Student's age |
| `Gender` | Student's gender |
| `Year_of_Study` | Year in their program |
| `Sleep_Hours` | Average hours of sleep per night |
| `Study_Hours` | Average hours of study per day |
| `Stress_Level` | Self-reported stress (1–10) |
| `Burnout_Level` | Self-reported burnout (1–10) |
| `Mental_Health_Score` | Overall mental health score |
| `Sought_Help` | Did they seek mental health help? (Yes/No) |

---

## Let's Load Everything

We just need the tidyverse and our dataset. If you haven't installed it yet, uncomment that first line and run it once.

```{r load-libraries}

library(tidyverse)
```

```{r load-data}
# Make sure the CSV is in the same folder as this .Rmd file!
df <- read_csv("student_mental_health_burnout.csv")

glimpse(df)
```
```{r clean-data}
# Dropping missing values and making sure categories are treated as factors
df <- df |>
  drop_na() |>
  mutate(
    gender       = as.factor(gender),
    academic_year = as.factor(academic_year),
    risk_level   = as.factor(risk_level)
  )

nrow(df)
```

## Plot 1: How Burned Out Are Students?

Let's start simple — just how bad is the burnout situation? A histogram gives us a quick look at how burnout scores are spread across all students.
```{r histogram-burnout}
ggplot(df, aes(x = burnout_score)) +
  geom_histogram(
    binwidth = 1,
    fill     = "#E76F51",
    color    = "white",
    alpha    = 0.85
  ) +
  labs(
    title    = "How Burned Out Are Students?",
    subtitle = "Burnout scored on a continuous scale",
    x        = "Burnout Level",
    y        = "Number of Students",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)
```

The burnout scores here are continuous — a score closer to 0 means little 
to no burnout, while anything pushing toward 10 is pretty severe. Most 
students are clustered on the lower end, which sounds good, but that big 
bar near 0 might just mean a lot of students haven't hit their breaking 
point *yet*.

---

## Plot 2: Does Burnout Differ by Gender?

Next let's see if burnout looks different across genders. A box plot is great for this. It'll show the middle range of scores and flags anyone who's way above or below average.

A taller box means scores are more spread out within that group. Compare the middle lines to see who's feeling it more.
```{r boxplot-gender}
ggplot(df, aes(x = gender, y = burnout_score, fill = gender)) +
  geom_boxplot(alpha = 0.7, outlier.color = "firebrick") +
  labs(
    title    = "Burnout by Gender",
    subtitle = "The line in the middle of each box is the median",
    x        = "Gender",
    y        = "Burnout Level (1–10)",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```


All three groups have a pretty similar median burnout score — so gender 
doesn't seem to make a huge difference for most students. But those red 
dots up top are worth noticing. Every gender has outliers pushing toward 
10, meaning there's a small group of seriously burned out students across 
the board.

"Other" likely includes non-binary and genderqueer students, or anyone 
who didn't identify with the Male/Female options on the survey.

## Plot 3: Do Stressed Students Burn Out More?

This one's pretty intuitive, if you're stressed, are you also more burned out? A scatter plot with a trend line helps us see if there's actually a pattern there.

If that line slopes upward, yeah — more stress really does mean more burnout.

```{r scatter-stress-burnout}
ggplot(df, aes(x = stress_level, y = burnout_score)) +
  geom_point(
    aes(color = gender),
    alpha = 0.5,
    size  = 2
  ) +
  geom_smooth(
    method = "lm",
    se     = TRUE,
    color  = "black",
    linewidth = 1
  ) +
  labs(
    title    = "Stress vs. Burnout",
    subtitle = "Each dot is one student — the black line is the overall trend",
    x        = "Stress Level (1–10)",
    y        = "Burnout Level (1–10)",
    color    = "Gender",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 13)
```


Yeah, pretty much what you'd expect, the trend line goes up steadily, 
meaning students with higher stress levels also tend to have higher burnout 
scores. The dots are packed pretty tight because there's a million rows of 
data, but the direction is clear. All three genders follow the same pattern 
too, which matches what we saw in the box plot.

---

## Plot 4: Does Burnout Get Worse Over the Years?

You'd think burnout piles up the longer you're in school. Let's check. We'll calculate the average burnout per year first, then plot it.

If the bars keep climbing, senior year is rough. If they're flat the burnout hits early and stays there.
```{r bar-year}
burnout_by_year <- df |>
  group_by(academic_year) |>
  summarise(
    Mean_Burnout = mean(burnout_score, na.rm = TRUE),
    SE           = sd(burnout_score, na.rm = TRUE) / sqrt(n())
  )

ggplot(burnout_by_year, aes(x = academic_year, y = Mean_Burnout, fill = academic_year)) +
  geom_bar(stat = "identity", alpha = 0.85, width = 0.6) +
  geom_errorbar(
    aes(ymin = Mean_Burnout - SE, ymax = Mean_Burnout + SE),
    width = 0.2, linewidth = 0.8
  ) +
  labs(
    title    = "Average Burnout by Year of Study",
    subtitle = "Error bars show ± 1 standard error",
    x        = "Year of Study",
    y        = "Mean Burnout Level",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```


Honestly, not what you'd expect! Burnout stays almost completely flat 
across all four years — first years and fourth years are equally burned 
out on average. With a million students in the dataset the error bars are 
tiny, so this isn't just noise. It seems like burnout hits early and just 
stays there rather than building up over time.
---

## Plot 5: Does Sleep Help? And Does Asking for Help Matter?

Last one. We're looking at whether more sleep means better mental health — and splitting it by whether the student sought help or not. `facet_wrap()` gives us two side-by-side panels so the comparison is easy.

If the slope is steeper for students who sought help, it suggests that sleep and professional support together make a real difference.
```{r facet-sleep-mental-health}
ggplot(df, aes(x = sleep_hours, y = mental_health_index, color = risk_level)) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ risk_level, labeller = label_both) +
  labs(
    title    = "Sleep Hours vs. Mental Health Score",
    subtitle = "Split by student dropout risk level",
    x        = "Sleep Hours per Night",
    y        = "Mental Health Score",
    color    = "Risk Level",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_color_manual(values = c("Low" = "#2A9D8F", "Medium" = "#E9C46A", "High" = "#E76F51")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The High risk panel is the most readable here — mental health scores are 
lower and more spread out for students flagged as high dropout risk, which 
makes sense. The Low and Medium panels are pretty packed since most students 
fall into those categories. The big takeaway is that high risk students have 
noticeably lower mental health scores overall, regardless of how much they sleep.

## Plot 6: Does Financial Stress Lead to Burnout?

Money stress is a real thing for students — tuition, rent, food. Let's see if it actually shows up in burnout scores.
```{r scatter-financial-burnout}
ggplot(df, aes(x = financial_stress, y = burnout_score)) +
  geom_point(
    aes(color = risk_level),
    alpha = 0.3,
    size  = 1.2
  ) +
  geom_smooth(
    method = "lm",
    se     = TRUE,
    color  = "black",
    linewidth = 1
  ) +
  labs(
    title    = "Financial Stress vs. Burnout",
    subtitle = "Each dot is one student — colored by dropout risk level",
    x        = "Financial Stress Score",
    y        = "Burnout Score",
    color    = "Risk Level",
    caption  = "Source: Kaggle – Student Mental Health & Burnout"
  ) +
  scale_color_manual(values = c("Low" = "#2A9D8F", "Medium" = "#E9C46A", "High" = "#E76F51")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

This one's pretty telling. The three risk levels naturally separate into 
bands — high risk students sit at the top with higher burnout scores, low 
risk at the bottom. The trend line slopes upward, so more financial stress 
does push burnout higher. But notice how the bands stretch across the full 
range of financial stress scores — students are dealing with money stress 
regardless of their risk level. It's just that high risk students are 
already burned out on top of it.

## Wrapping Up

So what did we actually find? A few things stand out across all six plots:

- Burnout is mostly low for most students, but there's always a subset 
who are really struggling — every plot showed that pattern
- Stress and burnout move together pretty closely — no surprise there
- Burnout doesn't get worse over the years, it just kind of stays flat 
from day one
- High dropout risk students have both lower mental health scores and 
higher burnout across the board
- Financial stress nudges burnout up, but students are dealing with money 
stress no matter what risk level they're in

As for ggplot2 — here's a quick recap of everything we used:

| Plot | Function | What it showed |
|---|---|---|
| Histogram | `geom_histogram()` | Overall spread of burnout scores |
| Box plot | `geom_boxplot()` | Burnout differences across genders |
| Scatter + trend | `geom_point()` + `geom_smooth()` | Stress vs. burnout |
| Bar chart | `geom_bar()` | Average burnout by year of study |
| Faceted scatter | `facet_wrap()` | Sleep vs. mental health by risk level |
| Scatter + trend | `geom_point()` + `geom_smooth()` | Financial stress vs. burnout |

---

*Dataset: [Kaggle – Student Mental Health and Burnout](https://www.kaggle.com/datasets/sharmajicoder/student-mental-health-and-burnout)*
---



