---
title: "TidyVerse Vignette: Exploring Netflix Data with dplyr and ggplot2"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This vignette demonstrates how to use core **TidyVerse** packages — specifically `dplyr` for data manipulation and `ggplot2` for visualization — using the [Netflix Movies and TV Shows dataset](https://www.kaggle.com/datasets/shivamb/netflix-shows) from Kaggle.

The dataset contains information about movies and TV shows available on Netflix as of mid-2021, including title, type, director, cast, country, date added, release year, rating, and duration.

---

## Load Libraries

```{r libraries}
library(tidyverse)  # loads dplyr, ggplot2, tidyr, stringr, and more
```

---

## Load the Data

We read in the CSV directly. You can download `netflix_titles.csv` from [Kaggle here](https://www.kaggle.com/datasets/shivamb/netflix-shows).

```{r load-data}
netflix <- read_csv("netflix_titles.csv")

glimpse(netflix)
```

---

## Data Cleaning with `dplyr`

### 1. Filter out rows with missing ratings

```{r filter}
netflix_clean <- netflix |>
  filter(!is.na(rating), !is.na(country))

nrow(netflix_clean)
```

### 2. Select only the columns we need

```{r select}
netflix_slim <- netflix_clean |>
  select(type, title, country, release_year, rating, duration, listed_in)

head(netflix_slim)
```

### 3. Mutate: Extract the primary country

Many entries list multiple countries. We use `mutate()` and `str_extract()` to grab just the first country listed.

```{r mutate}
netflix_slim <- netflix_slim |>
  mutate(primary_country = str_extract(country, "^[^,]+"))

head(netflix_slim |> select(country, primary_country))
```

---

## Summarizing Data with `dplyr`

### Content type breakdown

```{r count-type}
netflix_slim |>
  count(type, sort = TRUE)
```

### Top 10 countries by number of titles

```{r top-countries}
top_countries <- netflix_slim |>
  count(primary_country, sort = TRUE) |>
  slice_head(n = 10)

top_countries
```

### Average release year by content type

```{r avg-year}
netflix_slim |>
  group_by(type) |>
  summarise(
    avg_release_year = mean(release_year, na.rm = TRUE),
    count = n()
  )
```

---

## Visualization with `ggplot2`

### Plot 1: Movies vs TV Shows

```{r plot-type, fig.width=6, fig.height=4}
netflix_slim |>
  count(type) |>
  ggplot(aes(x = type, y = n, fill = type)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Netflix Content: Movies vs TV Shows",
    x = "Type",
    y = "Count"
  ) +
  theme_minimal()
```

### Plot 2: Top 10 Countries by Number of Titles

```{r plot-countries, fig.width=8, fig.height=5}
top_countries |>
  ggplot(aes(x = reorder(primary_country, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#a8d8ea", high = "#e84545") +
  labs(
    title = "Top 10 Countries on Netflix by Number of Titles",
    x = "Country",
    y = "Number of Titles"
  ) +
  theme_minimal()
```


