---
title: "LongLin Extension"
author: "Zihao Yu"
format: html
editor: visual
---

## Introduction

This vignette shows how to use Tidyverse packages like `dplyr` and `ggplot2` to organize data and display findings in a chart. The chosen data source named `global grocery inflation` is from Kaggle.

Source: https://www.kaggle.com/datasets/waddahali/global-grocery-inflation-20252026/data?select=breakfast+basket.csv

## Reading in the data with Tidyverse

We read in the data from Kaggle by first uploading the data to github and then using the `read_csv` method within the `tidyverse` library on a raw github link. Then we use the `glimpse` method within `dplyr` to visualize the columns within the data set.

```{r}
library(tidyverse)

url <- "https://raw.githubusercontent.com/longflin/DATA-607/refs/heads/main/TIDYVERSE2026/GlobalGroceryInflation.csv"

grocery_inflation <- read_csv(
  file = url,
  show_col_types = FALSE,
  progress = FALSE
)

glimpse(grocery_inflation)

```

Next, we use `select` within `dplyr` to select specific columns of interest. For this vignette, we will focus on the price of milk in target cities such as `New York`, `Los Angeles`, `London`, `Paris`, `Beijing`, and `Tokyo`. In order to do this, we use `filter` for `Item` that is equal to `Milk (1 Liter)` and for `City` that are in the `target_cities` variable.

```{r}
target_cities <- c("New York", "Los Angeles", "London", "Paris", "Beijing", "Tokyo")

milk_inflation <- grocery_inflation %>%
  select(City, Country, Item, Month, YoY_Inflation_Estimate_Pct) %>%
  filter(Item == "Milk (1 Liter)") %>%
  filter(City %in% target_cities)

head(milk_inflation)
```

Here, we use `summarize` to create a new column named `Average_YoY` to grab the average of all the months for each city.

```{r}
milk_inflation_group <- milk_inflation %>%
  group_by(City) %>% 
  summarize(
    Average_YoY = mean(YoY_Inflation_Estimate_Pct, na.rm = TRUE)
  )

head(milk_inflation_group)
```

In order to visualize the data, we use `ggplot2` to create a bar chart to display the results.

```{r}
library(ggplot2)

ggplot(milk_inflation_group, aes(x = reorder(City, Average_YoY), 
                                       y = Average_YoY, 
                                       fill = City)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(Average_YoY, 1)), 
            hjust = -0.2, 
            size = 3.5) + 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  labs(
    title = "Average Milk Inflation by City for Oct 2025 to March 2026",
    x = "City",
    y = "Average Year over Year Inflation (%)"
  ) +
  theme_minimal()
```

## Conclusion

Tidyverse is a very powerful tool for managing and transforming data. It allows you to create easy to read charts created from a data source of your choice. In this example, packages such as `dplyr` and `ggplot2`, made it very easy to manipulate the data and display the results.

### Tidyverse extend

Let's use lubridate and create a time plot of grocery prices.

```{r}
nyc <- grocery_inflation |> filter(City == "New York")

#convert the column to year/month (YM)
nyc <- nyc %>%
  mutate(Month = ym(Month))
```

### Creating a time plot

```{r}
#use geom_line + geom_point to include both lines and points
nyc |> ggplot(aes(x = Month, y = Price_USD, color = Item)) +
  geom_line() +
  geom_point(shape = 5) + #changing the shape of the point with "shape = 5"
  labs(title = "Grocery Prices Over a Six-Month Period", x = "Month (2025-2026)",
       y = "Price in USD") + 
  theme_minimal() #changing the background
```

### Mapping prices

ggplot has a mapping feature. However, you need coordinates, not just city names. We can use the maps package to join coordinates to the city name.

```{r}
library(maps)

data(us.cities)

#each city ends in the state, so to join we need to delete 3 characters on the end 
#of each city
us.cities$name <- str_sub(us.cities$name, end = -4)

#filter for only cities in the USA, eggs only

grocery_us <- grocery_inflation |> 
  filter(ISO_Country_Code == "USA") |> 
  filter(Item_Key == "Eggs_12") |> 
  filter(Month_Name == "March 2026")

#left join
cities_joined <- left_join(grocery_us, us.cities, join_by(City == name))



```

Plotting a map using the SF package:

```{r}
library(sf)

#use the SF Package to classify the coordinates
states_sf <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

cities_sf <- cities_joined |> 
  drop_na(long, lat) |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)


#the vibe is very Vegas in the 80s
ggplot() +
  #change the map colors
    geom_sf(data = states_sf, fill = "pink", color = "orange") +
    geom_sf(data = cities_sf, aes(color = Price_USD), size = 8) + 
  #ensure the colors get darker as the price gets higher
  scale_color_gradient(low = "#ADD8E6", high = "#000080") +
  #remove lines/background
    theme_void() + 
    labs(color = "Egg Price in Dollars", title = "Price of a Dozen Eggs in March 2026")

```

### Histogram of egg prices in March 2025 across the world

Are they normally distributed?

```{r}
grocery_inflation |> 
  #filter for one product in one month
  filter(Item_Key == "Eggs_12") |> 
  filter(Month_Name == "March 2026") |>
  #specify only x for histograms
  ggplot(aes(x = Price_USD)) +
  #specify binwidth or bins (or leave it blank and let ggpplot pick)
  geom_histogram(bins = 40)


#breakfast basket (price of a whole breakfast?), with different regions in different colors

grocery_inflation |> 
  filter(Item_Key == "Eggs_12") |> 
  filter(Month_Name == "March 2026") |>
  #specify fill for color
  ggplot(aes(x = Breakfast_Basket_USD, fill = Region)) +
  geom_histogram() +
  #remove the legend
  theme(legend.position = "none")
```

### Scatterplots

Plotting the population against the price of eggs:

```{r}

grocery_inflation |> 
  #population in millions instead of the actual number
  mutate(pop_est_mils = Population_Estimate/1000000) |>
  filter(Item_Key == "Eggs_12") |> 
  filter(Month_Name == "March 2026") |>
  #plotting population vs. price, breaking out color by country
  ggplot(aes(x = pop_est_mils, y = Price_USD, color = Country)) +
  #changing the shape of the points
  geom_point(shape = 3) + 
  #removing legend (othrwise it's overwhelming)
  theme(legend.position = "none") +
  #add titles
  labs(title = "Population vs. Egg Price", x = "City Population in Millions", y = "Egg Price in USD")

```

As population goes up, it looks like egg prices go down.

------------------------------------------------------------------------

# Extension by Zihao Yu

This section extends the original vignette by adding an additional analysis of the dataset.
It is intended to provide another example of how tidyverse functions can be used to summarize and visualize grocery inflation data.

```{r}
# Compare breakfast basket prices across the target cities in March 2026
breakfast_summary <- 
  grocery_inflation |>
  filter(City %in% target_cities, 
         Month_Name == "March 2026"
         ) |>
  select(City, Breakfast_Basket_USD
         ) |>
  distinct() |>
  arrange(desc(Breakfast_Basket_USD))

# Display the summary table
breakfast_summary
```

```{r}
# Create a bar chart of breakfast basket prices in March 2026
ggplot(breakfast_summary,
       aes(x = reorder(City, Breakfast_Basket_USD), 
           y = Breakfast_Basket_USD,)
) +
  geom_col(width = 0.7, fill = 'lightpink') +
  geom_text(aes(label = Breakfast_Basket_USD), hjust = -0.1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Breakfast Basket Prices in March 2026",
    x = "City",
    y = "Breakfast Basket Price (USD)"
) +
  theme_minimal()
```

This section compares breakfast basket prices across the target cities in March 2026 using both a summary table and a bar chart. The results show that New York has the highest breakfast basket price, while Tokyo has the lowest among the selected cities.

```{r}
# Calculate the price gap between the highest and lowest breakfast basket prices
price_gap <-
  breakfast_summary %>%
  summarise(
    highest_price = max(Breakfast_Basket_USD),
    lowest_price = min(Breakfast_Basket_USD),
    price_gap = highest_price - lowest_price
)

# Display the price gap summary
price_gap
```

This table shows the difference between the highest and lowest breakfast basket prices in March 2026. The price gap is 6.26 USD, indicating a noticeable variation in breakfast basket costs across the selected cities.

This extension adds additional analysis of breakfast basket prices across the target cities in March 2026. The summary table, bar chart, and price gap calculation provide a clearer comparison of city-level differences and show another way tidyverse can be used for summarization and visualization.