---
title: "TidyVerse CREATE assignment"
author: "Zihao Yu"
date: "2026-04-4"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This vignette uses the iris dataset to demonstrate basic data summarization with dplyr and visualization with ggplot2.

```{r}
library(tidyverse)

#Load the data
data(iris)

## Display the first 6 rows of the dataset
head(iris)
```

------------------------------------------------------------------------

This section summarizes petal length and petal width for each iris species. Comparing these statistics helps show how petal measurements differ across species.

```{r}
# Group the data by species
# Calculate summary statistics for petal length and petal width
petal_summary <- 
  iris %>%
  group_by(Species) %>%
  summarise(
    mean_petal_length = mean(Petal.Length),
    median_petal_length = median(Petal.Length),
    min_petal_length = min(Petal.Length),
    max_petal_length = max(Petal.Length),
    sd_petal_length = sd(Petal.Length),
    mean_petal_width = mean(Petal.Width),
    median_petal_width = median(Petal.Width),
    min_petal_width = min(Petal.Width),
    max_petal_width = max(Petal.Width),
    sd_petal_width = sd(Petal.Width)
)

# Display the summary table
petal_summary
```

--

This section summarizes sepal length and sepal width for each iris species. These results help compare sepal characteristics among the three species.

```{r}
# Calculate summary statistics for sepal length and sepal width
sepal_summary <- 
  iris %>%
  group_by(Species) %>%
  summarise(
    mean_sepal_length = mean(Sepal.Length),
    median_sepal_length = median(Sepal.Length),
    min_sepal_length = min(Sepal.Length),
    max_sepal_length = max(Sepal.Length),
    sd_sepal_length = sd(Sepal.Length),
    mean_sepal_width = mean(Sepal.Width),
    median_sepal_width = median(Sepal.Width),
    min_sepal_width = min(Sepal.Width),
    max_sepal_width = max(Sepal.Width),
    sd_sepal_width = sd(Sepal.Width)
  )

# Display the sepal summary table
sepal_summary
```

------------------------------------------------------------------------

This section uses boxplots to compare all four iris measurements across species. It provides a visual summary of the distribution of each variable.

```{r}
# Reshape the iris dataset into long format
# Create one combined boxplot for all four measurements by species
iris_long <- 
  iris %>%
  pivot_longer(
    cols = c(Sepal.Length, 
             Sepal.Width,
             Petal.Length, 
             Petal.Width),
    names_to = "Measurement",
    values_to = "Value"
)

ggplot(
  iris_long, 
  aes(x = Species, y = Value, fill = Species)
) +
  geom_boxplot() +
  facet_wrap(~ Measurement, scales = "free") +
  labs(
    title = "Comparison of Iris Measurements by Species",
    x = "Species",
    y = "Value"
)
```

------------------------------------------------------------------------

This scatter plot shows the relationship between sepal length and petal length. The three species form distinct clusters, and setosa is different from the other two species.

```{r}
# Create a scatter plot
ggplot(
  iris, 
  aes(x = Sepal.Length, y = Petal.Length, color = Species)
) +
  geom_point( size = 2) +
  labs(
    title = "Sepal Length vs Petal Length by Species",
    x = "Sepal Length",
    y = "Petal Length"
)
```

## Conclusion

This vignette demonstrated how dplyr can be used to summarize iris measurements by species and how ggplot2 can be used to visualize differences among species. Overall, the three iris species show differences, especially in petal measurements.

## Extension by Madina Kudanova: Petal Length-to-Width Ratio

This extension creates a new variable called `petal_ratio`, which compares petal length to petal width. This helps identify whether some species have longer, narrower petals.

```{r}
iris_ratio <- 
  iris %>%
  mutate(petal_ratio = Petal.Length / Petal.Width) %>%
  group_by(Species) %>%
  summarise(
    mean_petal_ratio = mean(petal_ratio),
    median_petal_ratio = median(petal_ratio),
    count = n()
  )

iris_ratio
```

The summary shows how the petal ratio differs across species.

```{r}
ggplot(iris_ratio, aes(x = Species, y = mean_petal_ratio, fill = Species)) +
  geom_col() +
  labs(
    title = "Average Petal Length-to-Width Ratio by Species",
    x = "Species",
    y = "Average Petal Ratio"
  )
```

## Extension by Emily El Mouaquite: Forcats and Ggplot

Using forcats and ggplot, the below bar chart displays the count of each species of iris in the data, ordered by how many times they appear. In this case, there are 50 observations of each species so it does not make any difference in the appearance of the chart, but is included in the code through the usage of the forcats function fct_infreq().

```{r}
iris %>%
  mutate(Species = fct_infreq(Species)) %>%
  ggplot(aes(x = Species, fill = Species)) +
  geom_bar() +
  labs(title = "Count of Observations by Species")
```