---
title: "TidyVerse Vignette: purrr and dplyr - EXTEND"
author: "Ciara Bonnett, David Chen"
format: html
editor: visual
---

## Introduction

This vignette demonstrated how to use the `dplyr` and `purrr` packages within the TidyVerse. We are using the built-in `msleep` dataset, which contains information about the sleep patterns of mammals. This collaborative example is designed to show how we can transform raw data into a clean, summarized format.

## Capability 1: Data Manipulation with dplyr

Here I am using`dplyr` to isolate herbivores. We calculate a new column for awake time and sort the results to see which heribivores sleep the most. This demonstrates the power of piping functions like `filter()`, `mutate(`, and `select()` together.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Cleaning and transforming the mammal sleep data
herbivore_sleep <- msleep %>%
  filter(vore == "herbi") %>%
  mutate(awake_time = 24 - sleep_total) %>%
  select(name, sleep_total, awake_time, brainwt) %>%
  arrange(desc(sleep_total))
  
# Display the first few rows
knitr::kable(head(herbivore_sleep), caption = "Top Sleeping Herbivores")
```

## Capability 2: Functional Programming with purrr

The `purrr` package is used here to apply the `mean` function across every numeric column in the dataset at once. This is much more efficient than writing individual lines of code for every column. We use `map_df` to ensure the output remains in a clean data frame format.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Use ggplot2::msleep to ensure R always sees it as the dataset
ggplot2::msleep %>%
  dplyr::select(where(is.numeric)) %>%
  purrr::map_df(~mean(.x, na.rm = TRUE)) %>%
  knitr::kable(caption = "Dataset Averages using purrr")
```

## Conclusion

This example shows how `dplyr` pipes (`%>%`) allows us to clean and transform data efficiently before passing it directly into a visualization.

## AI Transcript

Phase 1: Dataset Selection and Reasoning

User Inquiry: "So what dataset from kaggle do I need to use?" AI Response & Logic: The AI suggested that while Kaggle is an option, using a built-in TidyVerse dataset like msleep is superior for a "collaborative vignette" because it ensures the code will run for all classmates instantly without requiring them to download external files or handle broken file paths. Decision: The user chose to stay with the built-in msleep dataset for reliability and ease of extension by peers.

Phase 2: Troubleshooting the "Select" Error

User Inquiry: "Error in UseMethod('select') : no applicable method for 'select' applied to an object of class 'function'" Technical Logic: \* The Problem: The AI identified that the R environment was "masking" the TidyVerse select fuPhase 1: Dataset Selection and Reasoning

Phase 3: Final Polishing for Collaboration

User Inquiry: "Have I completed my code completely?" AI Response & Logic: The AI confirmed the code was functional but suggested adding "Annotations" (descriptive text). This is a requirement for a "vignette" to ensure that the classmate who has to "extend" the code later understands the logic.

## Extend

```{r}
print(msleep)
```

```{r}
colnames(msleep)
```

### Using a Basic "for" Loop

```{r}
for (i in 2:5) {
  col_name <- colnames(msleep)[i]
  unique_vals <- unique(msleep[[i]])
  
  cat("Unique values in column", col_name, ":\n")
  print(unique_vals)
  cat("---------------------------\n")
}
```

### Average Total Sleep Time by Conservation Status

```{r}
# Filer NAs and calculate means
sleep_summary <- msleep %>%
  filter(!is.na(conservation)) %>%
  group_by(conservation) %>%
  summarise(avg_sleep = mean(sleep_total)) %>%
  arrange(desc(avg_sleep))

# Create plot
ggplot(sleep_summary, aes(x = reorder(conservation, -avg_sleep), y = avg_sleep, fill = conservation)) +
  geom_col() +
  geom_text(aes(label = round(avg_sleep, 1)), vjust = -0.5) + # Add labels on top
  labs(title = "Average Total Sleep Time by Conservation Status",
       x = "Conservation Status",
       y = "Average Sleep (Hours)") +
  theme_minimal() +
  guides(fill = "none")
```

### Count of Observations per Conservation Status

```{r}
#
msleep_counts <- msleep %>%
  count(conservation) %>%
  arrange(desc(n))

print(msleep_counts)
```

```{r}
ggplot(msleep, aes(x = conservation, fill = conservation)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-0.5) + # Adds the count labels
  theme_minimal() +
  labs(title = "Count of Observations per Conservation Status",
       x = "Conservation Status",
       y = "Number of Animals")
```

### Filter for "en" and "nt", then create a rank index within each group
```{r}
 
line_plot_data <- msleep %>%
  filter(conservation %in% c("en", "nt")) %>%
  arrange(conservation, sleep_total) %>%
  group_by(conservation) %>%
  mutate(obs_id = row_number())
print(line_plot_data[,c(1,5)])

ggplot(line_plot_data, aes(x = obs_id, y = sleep_total, color = conservation)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Total Sleep Profile: Endangered vs. Near Threatened",
       subtitle = "Animals sorted from shortest to longest sleep duration",
       x = "Observation Rank",
       y = "Total Sleep (Hours)",
       color = "Status") +
  theme_minimal() +
  scale_color_manual(values = c("en" = "#1f77b4", "nt" = "#ff7f0e"))
```
The total sleep profile suggests that Endangered status does not dictate a specific sleep duration; rather, it reflects a group of animals with diverse biological needs that are currently under environmental pressure. Near Threatened species show a "tighter" profile, suggesting that their physiological sleep requirements may be more uniform across the sampled species. This comparison highlights that while sleep is a biological necessity, it is heavily influenced by the specific evolutionary traits of the animal rather than their conservation status alone.
